---
title: 'Team KPIs: NFL Example'
output:
  html_document:
    df_print: paged
---


# 1. Packages

```{r warning = F, message = F}
# install.packages('nflplotR')
# install.packages('randomForest')
# install.packages('ggrepel')
# install.packages('ggimage')
# install.packages('gt')
# remotes::install_github("jthomasmock/gtExtras")

library(tidyverse)
library(modelr)
library(nflreadr)
library(nflplotR)
library(randomForest)
library(ggrepel)
library(ggimage)
library(gt)
library(gtExtras)
```



# 2. Importing data:
Let's use the load_pbp() function to load in play-by-play data from the last three seasons and filter for the regular seasons.

```{r warning = F, message = F}
pbp <- load_pbp(seasons = 2019:2021)

pbp_regular <- pbp %>%
                filter(season_type == "REG")

pbp_playoff <- pbp %>%
                filter(season_type == "POST")

pbp
```



# 3. Produce a table of outcomes:
Aggregate the data on a team level to produce a table that shows the outcomes of performance of each team: how many games they won, games they tied, games they lost, Points for, and Points against. 

```{r warning = F, message = F}
outcomes_regular <- pbp_regular %>%
  group_by(season, game_id, home_team) %>%
  summarize(
    home_win = if_else(sum(result) > 0, 1, 0),
    home_tie = if_else(sum(result) == 0, 1, 0),
    home_diff = last(result),
    home_pts_for = last(home_score),
    home_pts_against = last(away_score)
  ) %>%
  group_by(season, home_team) %>%
  summarize(
    home_games = n(),
    home_wins = sum(home_win),
    home_ties = sum(home_tie),
    home_diff = sum(home_diff),
    home_pts_for = sum(home_pts_for),
    home_pts_against = sum(home_pts_against)
  ) %>%
  ungroup() %>%
  left_join(
    # away games
    pbp_regular %>%
      group_by(season, game_id, away_team) %>%
      summarize(
        away_win = if_else(sum(result) < 0, 1, 0),
        away_tie = if_else(sum(result) == 0, 1, 0),
        away_diff = last(result)*-1,
        away_pts_for = last(away_score),
        away_pts_against = last(home_score)
      ) %>%
      group_by(season, away_team) %>%
      summarize(
        away_games = n(),
        away_wins = sum(away_win),
        away_ties = sum(away_tie),
        away_diff = sum(away_diff),
        away_pts_for = sum(away_pts_for),
        away_pts_against = sum(away_pts_against)
      ) %>%
      ungroup(),
    by = c("season", "home_team" = "away_team")
  ) %>%
  rename(team = "home_team") %>%
  mutate(
    games = home_games + away_games,
    wins = home_wins + away_wins,
    ties = home_ties + away_ties,
    losses = games - wins - ties,
    point_diff = home_diff + away_diff,
    points_for = home_pts_for + away_pts_for,
    points_against = home_pts_against + away_pts_against) %>%
  select(season, team, games, wins, losses, ties, point_diff, points_for, points_against)

outcomes_regular
```



```{r}
outcomes_playoff <- pbp_playoff %>%
  group_by(season, game_id, home_team) %>% 
  summarize(
    home_win = if_else(sum(result) > 0, 1, 0),
    home_tie = if_else(sum(result) == 0, 1, 0),
    home_diff = last(result),
    home_pts_for = last(home_score),
    home_pts_against = last(away_score)
  ) %>%
  group_by(season, home_team) %>%
  summarize(
    home_games = n(),
    home_wins = sum(home_win),
    home_ties = sum(home_tie),
    home_diff = sum(home_diff),
    home_pts_for = sum(home_pts_for),
    home_pts_against = sum(home_pts_against)
  ) %>%
  ungroup() %>%
  left_join(
    # away games
    pbp_playoff %>%
      group_by(season, game_id, away_team) %>%
      summarize(
        away_win = if_else(sum(result) < 0, 1, 0),
        away_tie = if_else(sum(result) == 0, 1, 0),
        away_diff = last(result)*-1,
        away_pts_for = last(away_score),
        away_pts_against = last(home_score)
      ) %>%
      group_by(season, away_team) %>%
      summarize(
        away_games = n(),
        away_wins = sum(away_win),
        away_ties = sum(away_tie),
        away_diff = sum(away_diff),
        away_pts_for = sum(away_pts_for),
        away_pts_against = sum(away_pts_against)
      ) %>%
      ungroup(),
    by = c("season", "home_team" = "away_team")
  ) %>%
  rename(team = "home_team") %>%
  mutate(
    games = home_games + away_games,
    wins = home_wins + away_wins,
    ties = home_ties + away_ties,
    losses = games - wins - ties,
    point_diff = home_diff + away_diff,
    points_for = home_pts_for + away_pts_for,
    points_against = home_pts_against + away_pts_against) %>%
  select(season, team, games, wins, losses, ties, point_diff, points_for, points_against)

outcomes_playoff
```



# 4. Produce a table of metrics
Our KPIs will capture the following: number of passes, number of rushes, number of special team plays, pass yards, rush yards, epa per pass, epa per rush, success per pass, success per rush, yards per pass, yards per rush, and special teams epa per play. This will also be captured both from an offensive and defensive standpoint. 

```{r warning = F, message = F}
KPIs <- pbp_regular %>% 
  filter(pass == 1 & !is.na(epa) | rush == 1 & !is.na(epa) | special == 1 & !is.na(epa)
    ) %>% 
  group_by(season, posteam) %>% 
    summarize(
      n_pass = sum(pass),
      n_rush = sum(rush),
      n_special = sum(special),
      pass_yards = sum(yards_gained*pass, na.rm = TRUE),
      rush_yards = sum(yards_gained*rush, na.rm = TRUE),
      epa_per_pass = sum(epa*pass)/n_pass,
      epa_per_rush = sum(epa*rush)/n_rush,
      success_per_pass = sum(pass*epa>0)/n_pass,
      success_per_rush = sum(rush*epa>0)/n_rush,
      y_per_pass = sum(yards_gained*pass, na.rm = TRUE)/n_pass,
      y_per_rush = sum(yards_gained*rush, na.rm = TRUE)/n_rush,
      st_epa_per_play = sum(epa*special)/n_special
    ) %>% 
  left_join(
    pbp_regular %>%
      filter(pass == 1 & !is.na(epa) | rush == 1 & !is.na(epa) | special == 1 & !is.na(epa)
    ) %>% 
  group_by(season, defteam) %>% 
    summarize(
      def_n_pass=sum(pass),
      def_n_rush=sum(rush),
      def_n_special = sum(special),
      def_pass_yards = sum(yards_gained * pass, na.rm = TRUE),
      def_rush_yards = sum(yards_gained * rush, na.rm = TRUE),
      def_epa_per_pass=sum(-epa*pass)/def_n_pass,
      def_epa_per_rush=sum(-epa*rush)/def_n_rush,
      def_success_per_pass=sum(pass*epa>0)/def_n_pass,
      def_success_per_rush=sum(rush*epa>0)/def_n_rush,
      def_y_per_pass = sum(yards_gained*pass, na.rm = TRUE)/def_n_pass,
      def_y_per_rush = sum(yards_gained*rush, na.rm = TRUE)/def_n_rush,
      def_st_epa_per_play = sum(-epa*special)/def_n_special
    ),
  by = c("season", "posteam" = "defteam")
  ) %>% 
  rename(team = "posteam") %>% 
  select(-n_pass, -n_rush, -n_special,-def_n_pass, -def_n_rush, -def_n_special)

KPIs %>% 
  arrange(-st_epa_per_play)
```



# 5. Joining the outcomes and metrics tables: 
Let df represent the joining of the two results and metrics tables. 

```{r warning = F, message = F}
# Create dataframe for season long outcomes and stats
regular <- outcomes_regular %>% 
            left_join(KPIs, by = c("season", "team"))

regular
```


```{r}
playoff <- outcomes_playoff %>%
            select(season, team) %>% 
            mutate(season_team = str_c(season, '-', team))



df <- regular %>%
        left_join(playoff, by = c("season", "team"))

df <- df %>% 
        mutate(playoff = ifelse(is.na(df$season_team), 0, 1)) %>% 
        select(-season_team)

df
```



Since the 2021 season has 17 games, while 2020 and 2019 have 16 games, we'll create a win percentage column:

```{r}
df <- df %>%
        mutate(Win_pct = wins / games)

df
```


KPI Averages for Teams that made the Playoffs vs. teams that did NOT make the Playoffs

```{r}
playoff_kpi_avgs <- df %>% 
                      group_by(playoff) %>%
                      dplyr::summarize(
                                      avg_y_per_pass = mean(y_per_pass, na.rm=TRUE),
                                      avg_y_per_rush = mean(y_per_rush, na.rm=TRUE),
                                      avg_epa_per_pass = mean(epa_per_pass, na.rm=TRUE),
                                      avg_epa_per_rush = mean(epa_per_rush, na.rm=TRUE),
                                      avg_st_epa_per_play = mean(st_epa_per_play, na.rm=TRUE),
                                      avg_def_y_per_pass = mean(def_y_per_pass, na.rm=TRUE),
                                      avg_def_y_per_rush = mean(def_y_per_rush, na.rm=TRUE),
                                      avg_def_epa_per_pass = mean(def_epa_per_pass, na.rm=TRUE),
                                      avg_def_epa_per_rush = mean(def_epa_per_rush, na.rm=TRUE),
                                      avg_def_st_epa_per_play = mean(def_st_epa_per_play, na.rm=TRUE)
                                      )

playoff_kpi_avgs_tbl <- as.data.frame(t(playoff_kpi_avgs)) %>%        # transpose the data to vertical
                        rename(Playoff_NO = "V1", Playoff_YES = "V2")    # rename the columns
                    

playoff_kpi_avgs_tbl <- playoff_kpi_avgs_tbl[-1,]                     # remove the first row, which is simply Playoff 0/1
        
playoff_kpi_avgs_tbl
```



# 6. Correlation Analysis to Identify Team KPIs:
Identify measures of performance that correlate significantly with results. Ideally, the set of KPIs can span these various phases of the game, both offensive and defensive.  

```{r}
df %>%
  select(Win_pct, y_per_rush, y_per_pass, epa_per_rush, epa_per_pass, st_epa_per_play, def_y_per_rush,
         def_y_per_pass, def_epa_per_rush, def_epa_per_pass, def_st_epa_per_play) %>%
  cor() %>%
  data.frame() %>%
  select(Win_pct) %>%
  arrange(-Win_pct)
```



# 7. Let's build a Multiple Linear Regression model that considers all relevant KPIs:
Then we will use step-wise backward selection to remove variables that are not significant or are counter-intuitive. 

```{r}
WinPct_LR_Model <- lm(data = df, Win_pct ~ y_per_rush + y_per_pass + epa_per_rush + epa_per_pass + st_epa_per_play +
                      def_y_per_rush + def_y_per_pass + def_epa_per_rush + def_epa_per_pass + def_st_epa_per_play)
summary(WinPct_LR_Model)
```



Based on above model results, it would make sense to choose the following KPIs:
Offensive: EPA per pass (passing), EPA per rush (rushing), special teams EPA per play (special team)
Defensive: Defensive EPA per pass (passing), Defensive EPA per rush (rushing)

Since the correlation for defensive special teams epa per play was counter-intuitive, we should consider not retaining it as a KPI. Part of what impacts this particular measure is the kicking game, and from a defensive standpoint, you have less control over this. 

The good news here is that all the variables selected for the model were significant. The Adjusted R-square was also high at 0.804. Thus, we can say that 80.40% of the variation in win percentage can be explained by the regression model. 

```{r}
WinPct_LR_Model <- lm(data = df, Win_pct ~ epa_per_pass + epa_per_rush + st_epa_per_play + def_epa_per_pass + def_epa_per_rush)
summary(WinPct_LR_Model)
```



# 8. Examining Actual Win Percentage vs. Expected Win Percentage
First, let's layer on the predicted and residual values from the model and store this as a new table called df_predict 

```{r}
df <- df %>%
        add_predictions(WinPct_LR_Model) %>%
        add_residuals(WinPct_LR_Model) %>%
        rename(x_LR_WinPct = pred, LR_Resid = resid)

df
```



# 9. Build a second model examine it's feature importance to determine most impactful KPIs and KPI cutoffs for Win Pct. 

```{r}
#make this example reproducible
set.seed(32)

#fit the random forest model
WinPct_RF_Model <- randomForest(formula = Win_pct ~ y_per_rush + y_per_pass + epa_per_rush + epa_per_pass + st_epa_per_play +
                                def_y_per_rush + def_y_per_pass + def_epa_per_rush + def_epa_per_pass + def_st_epa_per_play, 
                                data = df)
```



We can see that this RandomForest model lists epa_per_pass and def_epa_per_pass as the two most important features. This is in line with the regression model's outputs where the two largest coefficients correspond to these same two variables.

```{r}
RF_Feature_Importance <- as.data.frame(importance(WinPct_RF_Model))
RF_Feature_Importance %>% 
  arrange(-IncNodePurity)
```



Add Random Forest Predictions to DataFrame

```{r}
df <- df %>% 
        add_predictions(WinPct_RF_Model) %>%
        add_residuals(WinPct_RF_Model) %>% 
        rename(x_RF_WinPct = pred, RF_Resid = resid)
```



# 9. Visualizing NFL Team Data

First, we'll look to do some tabular representations of the data relating to team performance measures. In order to enhance the data visualization, team logos and colors will be used. In this step, we will join that data to our df table. 

```{r}
logos <- nflreadr::load_teams()

df <- df %>%
        inner_join(logos, by = c("team" = "team_abbr"))

df
```



Next, let's filter the data to just look at the 2021 season. 

```{r}
df_2021 <- df %>%
            filter(season == 2021)

df_2021
```



WIN PERCENTAGE

```{r warning = F, message = F}
df_2021 %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct) %>%
    mutate(Rank = rank(-Win_pct, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("Win_pct"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Win Pct. in 2021 Season"))
```



YARDS PER PASS

```{r}
y_per_pass <- merge(df_2021, playoff_kpi_avgs_tbl["avg_y_per_pass",])
y_per_pass %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, y_per_pass, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-y_per_pass, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("y_per_pass", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Yards per Pass in the 2021 Season"))
```



YARDS PER RUSH

```{r}
y_per_rush <- merge(df_2021, playoff_kpi_avgs_tbl["avg_y_per_rush",])
y_per_rush %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, y_per_rush, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-y_per_rush, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("y_per_rush", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Yards per Rush in the 2021 Season"))
```



EPA PER PASS

```{r warning = F, message = F}
epa_per_pass <- merge(df_2021, playoff_kpi_avgs_tbl["avg_epa_per_pass",])
epa_per_pass %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, epa_per_pass, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-epa_per_pass, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("epa_per_pass", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of EPA per Pass in the 2021 Season"))
```


EPA PER RUSH

```{r}
epa_per_rush <- merge(df_2021, playoff_kpi_avgs_tbl["avg_epa_per_rush",])
epa_per_rush %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, epa_per_rush, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-epa_per_rush, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("epa_per_rush", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of EPA per Rush in the 2021 Season"))
```



SPECIAL TEAMS EPA PER PLAY

```{r}
st_epa_per_play <- merge(df_2021, playoff_kpi_avgs_tbl["avg_st_epa_per_play",])
st_epa_per_play %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, st_epa_per_play, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-st_epa_per_play, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("st_epa_per_play", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Special Teams EPA per Play in the 2021 Season"))
```



Defensive YARDS PER PASS

```{r}
def_y_per_pass <- merge(df_2021, playoff_kpi_avgs_tbl["avg_def_y_per_pass",])
def_y_per_pass %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, def_y_per_pass, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(def_y_per_pass, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("def_y_per_pass", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Defensive Yards per Pass in the 2021 Season"))
```



Defensive YARDS PER RUSH

```{r}
def_y_per_rush <- merge(df_2021, playoff_kpi_avgs_tbl["avg_def_y_per_rush",])
def_y_per_rush %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, def_y_per_rush, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(def_y_per_rush, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("def_y_per_rush", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Defensive Yards per Rush in the 2021 Season"))
```



Defensive EPA PER PASS

```{r}
def_epa_per_pass <- merge(df_2021, playoff_kpi_avgs_tbl["avg_def_epa_per_pass",])
def_epa_per_pass %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, def_epa_per_pass, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-def_epa_per_pass, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("def_epa_per_pass", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Defensive EPA per Pass in the 2021 Season"))
```



Defensive EPA PER RUSH

```{r}
def_epa_per_rush <- merge(df_2021, playoff_kpi_avgs_tbl["avg_def_epa_per_rush",])
def_epa_per_rush %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, def_epa_per_rush, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-def_epa_per_rush, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("def_epa_per_rush", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Defensive EPA per Rush in the 2021 Season"))
```



Defensive SPECIAL TEAMS PER PLAY

```{r}
def_st_epa_per_play <- merge(df_2021, playoff_kpi_avgs_tbl["avg_def_st_epa_per_play",])
def_st_epa_per_play %>%
    select(team_wordmark, team, wins, losses, ties, Win_pct, def_st_epa_per_play, Playoff_NO, Playoff_YES) %>%
    mutate(Rank = rank(-def_st_epa_per_play, ties.method = "min")) %>%
    select(Rank, everything()) %>%
    arrange(-Rank) %>%
    gt::gt() %>%
    fmt_number(columns = c("def_st_epa_per_play", "Win_pct", "Playoff_NO", "Playoff_YES"), decimals = 3) %>%
    gt_img_rows(team_wordmark) %>%
    tab_style(style = list(cell_fill(color = "yellow")),
    locations = gt::cells_body(rows = team == "ATL")) %>%
    tab_header(title = md("Team Rankings of Defensive Special Teams EPA per Play in the 2021 Season"))
```



## Scatterplot visuals
When doing team evaluation, scatterplots can be a good way to show team performance in relation to the rest of the league by looking at the interaction between two variables. This can also help to classify teams according to their strengths/weaknesses.
Let's do a scatterplot of epa_per_pass vs. def_epa_per_pass with points being labeled by teams.

```{r}
 df_2021 %>%
  mutate(selected_team = ifelse(team == "ATL", 1, 0)) %>%
  ggplot(aes(x = epa_per_pass, y = def_epa_per_pass)) +
    geom_point(aes(alpha = selected_team, color = selected_team), size = 12, show.legend = F) +
    geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
    geom_hline(aes(yintercept = mean(def_epa_per_pass)), color = "red", linetype = "dashed") +
    geom_vline(aes(xintercept = mean(epa_per_pass)), color = "red", linetype = "dashed") +
    geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
    labs(x = "Offensive Pass EPA per play",
         y = "Defensive EPA per play",
         title = "2021 NFL Offensive and Defensive Pass EPA per Play") +
  theme_bw() +
  theme(aspect.ratio = 9 / 16,
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_colour_gradient(low = "white", high = "yellow")
```



Lastly, as opposed to splitting the scatterplot into four quadrants, it may be preferable to simply look at fitted relationships and teams that may be above or below the fitted line.  Let's produce a scatter plot with team logos using Win Percentage (y) and Expected Win Percentage (x). We'll also include a fitted line and highlight the team of interest. 

```{r}
 df_2021 %>%
  mutate(selected_team = ifelse(team == "ATL", 1, 0)) %>%
  ggplot(aes(x = Win_pct, y = x_LR_WinPct)) +
    geom_point(aes(alpha = selected_team, color = selected_team), size = 12, show.legend = F) +
    geom_smooth(method = lm, se = F) +
    geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
    labs(x = "Win Percentage",
         y = "Expected Win Percentage (Linear Regression)",
         title = "2021 NFL Expected Win Pct. vs Actual Win Pct. (Linear Regression)") +
  theme_bw() +
  theme(aspect.ratio = 9 / 16,
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_colour_gradient(low = "white", high = "yellow")
```



EPA per Pass vs. EPA per Rush

```{r}
 df_2021 %>%
  mutate(selected_team = ifelse(team == "ATL", 1, 0)) %>%
  ggplot(aes(x = epa_per_rush, y = epa_per_pass)) +
    geom_point(aes(alpha = selected_team, color = selected_team), size = 12, show.legend = F) +
    geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
    geom_vline(aes(xintercept = mean(epa_per_rush)), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = mean(epa_per_pass)), color = "red", linetype = "dashed") +
    geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
    geom_smooth(method = lm, se = F) +
    labs(x = "EPA per Rush",
         y = "EPA per Pass",
         title = "2021 NFL EPA per Pass vs. EPA per Rush") +
  theme_bw() +
  theme(aspect.ratio = 9 / 16,
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_colour_gradient(low = "white", high = "yellow")
```



EPA per Pass vs. Defensive EPA per Pass

```{r}
 df_2021 %>%
  mutate(selected_team = ifelse(team == "ATL", 1, 0)) %>%
  ggplot(aes(x = epa_per_pass, y = def_epa_per_pass)) +
    geom_point(aes(alpha = selected_team, color = selected_team), size = 12, show.legend = F) +
    geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
    geom_vline(aes(xintercept = mean(epa_per_pass)), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = mean(def_epa_per_pass)), color = "red", linetype = "dashed") +
    geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
    geom_smooth(method = lm, se = F) +
    labs(x = "EPA per Pass",
         y = "Defensive EPA per Pass",
         title = "2021 NFL EPA per Pass vs. Defensive EPA per Pass") +
  theme_bw() +
  theme(aspect.ratio = 9 / 16,
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_colour_gradient(low = "white", high = "yellow")
```



EPA per Rush vs. Defensive EPA per Rush

```{r}
 df_2021 %>%
  mutate(selected_team = ifelse(team == "ATL", 1, 0)) %>%
  ggplot(aes(x = epa_per_rush, y = def_epa_per_rush)) +
    geom_point(aes(alpha = selected_team, color = selected_team), size = 12, show.legend = F) +
    geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
    geom_vline(aes(xintercept = mean(epa_per_rush)), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = mean(def_epa_per_rush)), color = "red", linetype = "dashed") +
    geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
    labs(x = "EPA per Rush",
         y = "Defensive EPA per Rush",
         title = "2021 NFL EPA per Rush vs. Defensive EPA per Rush") +
  theme_bw() +
  theme(aspect.ratio = 9 / 16,
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_colour_gradient(low = "white", high = "yellow")
```



Special Teams EPA per Play vs. Defensive Special Teams EPA per Play 

```{r}
 df_2021 %>%
  mutate(selected_team = ifelse(team == "ATL", 1, 0)) %>%
  ggplot(aes(x = st_epa_per_play, y = def_st_epa_per_play)) +
    geom_point(aes(alpha = selected_team, color = selected_team), size = 12, show.legend = F) +
    geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
    geom_vline(aes(xintercept = mean(epa_per_rush)), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = mean(def_epa_per_rush)), color = "red", linetype = "dashed") +
    geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
    labs(x = "Special Teams EPA per Play",
         y = "Defensive Special Teams EPA per Play",
         title = "2021 NFL Special Teams EPA per Play vs. Defensive Special Teams EPA per Play") +
  theme_bw() +
  theme(aspect.ratio = 9 / 16,
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold")) +
  scale_colour_gradient(low = "white", high = "yellow")
```